---
title: "Your first R session"
output:
  html_document:
    toc: true
    theme: united
---


```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```


> ## Learning Objectives
>
> * Familiarize participants with R syntax
> * Understand the concepts of objects and assignment
> * Understand the concepts of vector and data types
> * Get exposed to a few functions



## Creating objects

```{r, purl=TRUE, echo=FALSE}
## Creating objects
```

Let's start by creating a simple object:

```{r, purl=FALSE}
x <- 10
x
```

We assigned to `x` the number 10. `<-` is the assignment operator. It assigns
values on the right to objects on the left. Mostly similar to `=` but not
always. Learn to use `<-` as it is good programming practice. Using `=` in place
of `<-` can lead to issues down the line. In RStudio, use the key combination
`Alt + -` to generate the assignment operator.

`=` should only be used to specify the values of arguments in functions for
instance `read.csv(file="data/some_data.csv")`.

Now that `x` exists in R memory, We can do things with it. For instance:

```{r, purl=FALSE}
x * 2
x + 5
x + x
```

or we can create new objects using `x`:

```{r, purl=FALSE}
y <- x + x + 5
```

Let's try something different:

```{r, purl=FALSE}
x <- c(2, 4, 6)
x
```

Two things:

- we overwrote the content of `x`
- `x` now contains 3 elements

Using the `[]`, we can access individual elements of this object:

```{r, purl=FALSE}
x[1]
x[2]
x[3]
```

---

### Challenge

Without typing this in R, can you guess the content of this vector?

```{r, purl=TRUE}
q <- c(x, x, 5)
```

What is the value of `y` at the end of this script?

```{r, purl=TRUE}
x <- 1
y <- x + 2
x <- 3
# y <- ???
```

---

Objects can be given any name such as `x`, `current_temperature`, or
`subject_id`. You want your object names to be explicit and not too long. They
cannot start with a number (`2x` is not valid but `x2` is). R is case sensitive
(e.g., `weight_kg` is different from `Weight_kg`). There are some names that
cannot be used because they represent the names of fundamental functions in R
(e.g., `if`, `else`, `for`, see
[here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Reserved.html)
for a complete list). In general, even if it's allowed, it's best to not use
other function names (e.g., `c`, `T`, `mean`, `data`, `df`, `weights`). In doubt
check the help to see if the name is already in use. It's also best to avoid
dots (`.`) within a variable name as in `my.dataset`. It is also recommended to
use nouns for variable names, and verbs for function names.


We can also use these objects with functions, for instance to compute the mean
and the standard deviation:

```{r, purl=FALSE}
x <- c(1, 3, 5, 6, 3, 5)
mean(x)
sd(x)
```

This is useful to print the value of the mean or the standard deviation, but we
can also save these values in their own variables:

```r
mean_x <- mean(x)
mean_x
```

The function `ls()` returns the objects that are currently in the memory of
your session.

The function `data()` allows you to load into memory datasets that are provided
as examples with R (or some packages). Let's load the `Nile` dataset that
provides the annual flow of the river Nile between 1871 and 1970.

```r
data(Nile)
```

Using `ls()` shows you that the function `data()` made the variable `Nile`
available to you.

Let's make an histogram of the values of the flows:

```r
hist(Nile)
```

---

### Challenge

The following: `abline(v=100, col="red")` would draw a vertical line on an
existing plot at the value 100 colored in red.

How would you add such a line to our histogram to show where the mean falls in
this distribution?

---

We can now save this plot in its own file:

```{r, eval=FALSE}
pdf(file="nile_flow.pdf")
hist(Nile)
abline(v=mean(Nile), col="red")
dev.off()
```
