---
title: "Vectors"
output:
  html_document:
    toc: true
    theme: united
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(results='hide', fig.path='img/r-lesson-')
```


## Vectors

Vectors are at the heart of how data are stored into R's memory. Almost
everything in R is stored as a vector. When we typed `x <- 10` we created a
vector of length 1. When we typed `x <- c(2, 4, 6)` we created a vector of
length 3. These vectors are of class `numeric`. Vectors can be of 6 different
classes (we'll mostly work with 4).

We have already seen how to create numeric vectors, but vectors can also contain
characters:

```{r, purl=FALSE}
mass <- c(50, 60, 65, 82)
mass
animals <- c("mouse", "rat", "dog", "elephant")
animals
```

### The different "classes" of vector

Vectors can hold numbers, characters, logical values and a special class that
characterizes R, factors.

* `"numeric"` is the general class for vectors that hold numbers (e.g., `c(1, 5,
  10)`)
* `"integer"` is the class for vectors for integers. To differentiate them from
  `numeric` we must add an `L` afterwards (e.g., `c(1L, 2L, 5L)`)
* `"character"` is the general class for vectors that hold text strings (e.g.,
  `c("blue", "red", "black")`)
* `"logical"` for holding `TRUE` and `FALSE` (boolean data type)

The other types of vectors are `"complex"` (for complex numbers) and `"raw"` a
special internal type that is not of use for the majority of users.

If you want to check the class of your vector, you can use the `class()` function.

```{r}
class(mass)
class(animals)
```

### How to check the length of your vector

```{r}
length(mass)
length(animals)
```


### How to add elements to your vector

You can add elements to your vector simply by using the `c()` function:

```{r, purl=FALSE}
mass <- c(mass, 90) # adding at the end
mass <- c(30, mass) # adding at the beginning
mass
```

### How to access elements of a vector?

They can be accessed by their indices:

```r
mass[1]
mass[c(2, 3)]
```

repeatitions are allowed:

```r
mass[c(2,3,2,4,1,2)]
```

### How to update/replace the value of a vector?

```r
mass[4] <- 122
```


### How to remove elements from a vector?

```r
mass[-2]          # doesn't affect the content of mass
mass[-c(1, 3, 5)]
```


### Sequences

`:` is a special function that creates numeric vectors of integer in increasing
or decreasing order, test `1:10` and `10:1` for instance. The function `seq()`
(for __seq__uence) can be used to create more complex patterns:

```{r, results='show', purl=FALSE}
seq(1, 10, by=2)
seq(5, 10, length.out=3)
seq(50, by=5, length.out=10)
seq(1, 8, by=3) # sequence stops to stay below upper limit
seq(1.1, 2, length.out=10)
```

### Repeating

```r
x <- rep(8, 4)
x
rep(1:3, 3)
```

### Operations on vectors

```r
x <- c(5, 10, 15)
x + 10
x + c(10, 15, 20)
x * 10
x * c(2, 4, 3)
```

Note that operations on vectors are elementwise.

### Recycling

R allows you to do operations on vectors of different lengths. The shorter
vector will be "recycled" (~ repeated) to match the length of the longer one:

```r
x <- c(5, 10, 15)
x + c(2, 4, 6, 8, 10, 12)     # no warning when it's a multiple
x + c(2, 4, 6, 8, 10, 12, 14) # warning
```

### Boolean operations and Filtering

```r
u <- c(1, 4, 2, 5, 6, 3, 7)
u[TRUE, FALSE, TRUE, TRUE, TRUE, FALSE, FALSE]
u < 3
u[u < 3]
u[u < 3 | u >= 4]
u[u > 5 & u < 1 ] ## nothing matches this condition
u[u > 5 & u < 8]
sum(u < 3)
```
